%! Author = matt_dumont
%! Date = 4/12/23

\color{ksldarkblue} \LARGE \textbf{Kōmanawa:}
\color{black} \large
\begin{enumerate}
    \item (verb) to spring, well up (of water)
    \item (verb) to spring, well up (of thoughts, ideas)
\end{enumerate}
\normalsize

Kōmanawa Solutions Limited (KSL) is a water resource consultancy and research company specialising in water resource investigation and modelling, environmental limit setting and water resource impact assessment. Our goal is to provide excellent science to facilitate the robust management of natural resources in our changing climate. Clients include New Zealand enterprises in the private sector, central and local government agencies and community groups.
Our vision
KSL delivers high quality science and research. We aspire to be at the forefront of creativity and innovation to address our increasingly complex water resource challenges; mō tatou, ā, mō kā uri ā muri ake nei (for us and our children after us).

\color{ksldarkblue} \LARGE \textbf{Our mission}
\color{black} \normalsize

Our mission is to develop solutions to the increasingly challenging water resource management issues we now face by providing a clear vision of the pathway from problem to solution. We work closely with our partners, communities, and stakeholders, deploying state-of-the-art scientific methods and building trust through knowledge and honest science communication.

\color{ksldarkblue} \LARGE \textbf{Limitations}
\color{black} \normalsize

Kōmanawa Solution Ltd (KSL) has prepared this Report in accordance with the usual care and thoroughness of the consulting profession for the use of \client{}.

This Report has been prepared in accordance with the scope of work and for the purpose outlined at the start of this report and is based on generally accepted practices and standards at the time it was prepared. No other warranty, expressed or implied, is made as to the professional advice included in this Report.

Where this Report indicates that information has been provided to KSL by third parties, KSL has made no independent verification of this information except as expressly stated in the Report. KSL assumes no liability for any inaccuracies in or omissions to that information.


This Report was prepared between \reportstartdate{} and \reportenddate{} and is based on the conditions encountered and information reviewed at the time of preparation. KSL disclaims responsibility for any changes that may have occurred after this time.

This Report should be read in full. No responsibility is accepted for use of any part of this Report in any other context or for any other purpose. This Report does not purport to give legal advice. Legal advice can only be given by qualified legal practitioners.

This Report has been prepared for the exclusive use of \client{} and their authorised agents. Except as required by law, no third party may use or rely on this Report unless otherwise agreed in writing by KSL.

To the extent permitted by law, KSL expressly disclaims and excludes liability for any loss, damage, cost or expenses suffered by any third party relating to or resulting from the use of, or reliance on, any information contained in this Report. KSL does not admit that any action, liability or claim may exist or be available to any third party.

\newpage

\color{ksldarkblue} \LARGE \textbf{Version Control}
\color{black} \normalsize

% add version control table
\input{./tables/version_control_table}

\color{ksldarkblue} \LARGE \textbf{Authors}
\color{black} \normalsize

\authors


% contributors empty file is ok here
\input{./custom_sections/contribuitors}

% box with suggested citation auto generated from doc info
\begin{breakawaybox}[label={box:sugcite}, title=Suggested Citation, flushleft title]
    \textbf{\authorcite{}},
    (\getYear{\reportenddate}). \textit{\reporttitle{}}.
    Report: \projcode{}.\nthreport{} prepared by Kōmanawa Solutions Ltd for \client.
\end{breakawaybox}

\newpage